---
title: "Lecture Code Notes"
author: "Maddie"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# remember to load stan in r package?


```


## Lecture 14: Model checking


```{r}

library(rstan)
library(tidyverse)

sppnames <- c( "afarensis","africanus","habilis","boisei","rudolfensis","ergaster","sapiens")
brainvolcc <- c( 438 , 452 , 612, 521, 752, 871, 1350 )
masskg <- c( 37.0 , 35.5 , 34.5 , 41.5 , 55.5 , 61.0 , 53.5 )
d <- data.frame( species=sppnames , brain=brainvolcc , mass=masskg )

# normalization

d$mass_std <- (d$mass - mean(d$mass))/sd(d$mass)
d$brain_std <- d$brain / max(d$brain)


# fit in stan 

write("// Stan model for parametric equations

data {
 int < lower = 1 > N; // sample size
 vector[N] Y; // response
 vector[N] X; // predictor, year
 vector[N] X_2; // predictor sq
}

parameters {
  real a; // intercept
  real bM; // slope
  real bM2; // slope 2
  real sigma; // error
}

model {
  vector[N] mu;
  mu = a + bM*X + bM2*X_2;
  Y ~ normal(mu, sigma);
  a ~ normal(0.5,1);
  bM ~ normal(0,10);
  bM2 ~ normal(0,10);
  sigma ~ normal(0,1);
}

generated quantities {
} 
",

"stan_ch7.2.stan")

stanc("stan_ch7.2.stan")

stan_data <- list(N = nrow(d), X=d$mass_std, Y=d$brain_std)

fit.ch7.1 <- rstan::stan(file="stan_ch7.1.stan", data=stan_data, chains=3, iter=1000)
fit.ch7.1

d$X_2 <- (d$mass_std)^2
stan_data2 <- list(N = nrow(d), X=d$mass_std, X_2=d$X_2, Y=d$brain_std)
fit.ch7.2 <-  rstan::stan(file="stan_ch7.2.stan", data=stan_data2, chains=3, iter=1000)
fit.ch7.2

mcmc_trace(fit.ch7.2)

df <- spread_draws(fit.ch7.2)

```



## Use rethinking for exploratory and prior predictive check

[insert code when library loads]

## Write height model in STAN

In English: "What is the mean height of adults in Dobe that are members of the !kung san?"

```{r}


write("//
      data{
      int < lower = 1 > N; // sample size
      vector[N] y; // response
      }
      
      parameters{
      real mu; 
      real < lower = 0 > sigma; // we know sigma wont be zero, and 50 is huge
      }
      
      model{
      y ~ normal(mu,sigma);
      mu ~ normal(178,20);
      sigma ~ uniform(0,50);
      }
      
      ")


```

Adding in weight:

Prior predictive check
```{r}

set.seed(100)
N <- 100 # 100 lines in plot of a and b
a <- rnorm(N, 178, 20)
b <- rnorm(N, 0, 10)
# can also try other distribution priors:
b <- rlnorm(N, 0, 1) # <- see plot is much more informative, maybe too informative.

# plot


```

Fit model using stan:
```{r}

write("//
      data{
      int < lower = 1 > N; // sample size
      vector[N] y; // response
      vector[N] x; // predictor
      }
      
      parameters{
      real alpha; 
      real < lower = 0 > beta; // slope
      real < lower = 0, upper = 50 > sigma; // error
      }
      
      model{
      vector[N] mu; // mu no longer a parameter
      mu = + beta*x; 
      y ~ normal(mu,sigma);
      alpha ~ normal(178,20);
      beta ~ lognormal(0,1)
      sigma ~ uniform(0,50);
      }
      generated quantities{
      
      }
      
      ",
      "stan_m4.3.stan")

stanc("stan_m4.3.stan") # compile

# add data, fit model:

d2$weight2 <- d2$weight - mean(d2$weight)
stan_data <- list(N= nrow(d2), x = d2$weight2, y = d2$height)

fit.m4.3 <- rstan::stan(file = "stan_m4.3.stan")

# do diagnostics: r hat, traceplot


# use tidybayes to make the plots


```


# Lecture 6 - 7
## Globe Tossing Model in Stan
- three chunks (sometimes 4)
- for `data` we need to define what kind, ie "int" for integer
- for `parameter` we also need to tell it what its storing, ie real for real number (can be a decimal), and define parameter space
- semi colon ends each line
- slashes are comments
- `model` needs to have likelihood and prior

```{r}

# this is writing a file out with all this stuff in it!
write("// Stan Model for Globe toss

data {
int < lower = 0 > N; // number of tosses,
int < lower = 0, upper = N > W;// times we land on water
}

parameters{
real < lower = 0, upper = 1 > p; // probability of water
}

model {
p ~ uniform(0,1); //prior
W ~ binomial(N,p); // likelihood
}
",

"stan_globe_toss.stan")

stanc("stan_globe_toss.stan") # compile Stan code in C++
```

