---
title: 'Assignment 2: 2M1 and 2M2'
author: "Madeline Berger"
date: "2026-01-26"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)




```

## 2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for *each of the following sets of observations*. In each case, assume a uniform prior for p.

(1) W, W, W
(2) W, W, W, L
(3) L, W, W, L, W, W, W

Steps to grid approximation:

-   define the grid, i.e. how many points to use in estimating the
    posterior, and make a list of the parameter values in the grid

-   compute the value of the prior at each parameter value in the grid

-   compute the likelihood at each parameter value

-    compute the standardized posterior at each parameter value, by
    multiplying the prior by the likelihood

-   finally standardize the posterior by dividing each value by the sum
    of all values

**W, W, W**

```{r}
# define the grid, uniform prior for p

p_grid <- seq(from = 0, to = 1, length.out = 20) # this can be any number, but sufficiently large to get a reasonable approximation of posterior 
# values of parameter p for which we want to assign a posterior to 
# for each value of p, what is the probability of that parameer given the data under the binomial distribution 


# define prior. uniform prior. 
prior <- rep(1,20) # multiply anything by itself, gives no information. it could be something else, but that might give you more information

#compute likelihood - density of binomial distribution.
likelihood <- dbinom(3, size = 3, prob = p_grid) # this is where the length of data and the results WWW come in. 3 W in 3 trials

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardixe Posterior 

posterior <- unstd.posterior / sum(unstd.posterior)


plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext( "Probability of Water, given WWW" )
```

**W, W, W, L**

```{r}

# define the grid, uniform prior for p

p_grid <- seq(from = 0, to = 1, length.out = 20) # 4 observations

# define prior 
prior <- rep(1,20)

#compute likelihood
likelihood <- dbinom(3, size = 4, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardixe Posterior 

posterior <- unstd.posterior / sum(unstd.posterior)


plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext( "4 points - W W W L" )




```

**L, W, W, L, W, W, W**

```{r}


# define the grid, uniform prior for p

p_grid <- seq(from = 0, to = 1, length.out = 20) # 7 observations

# define prior - not sure I did this right
prior <- rep(1,20)

#compute likelihood
likelihood <- dbinom(4, size = 7, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardixe Posterior 

posterior <- unstd.posterior / sum(unstd.posterior)


plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext( "7 points - L, W, W, L, W, W, W" )



```

## 2M2. Now assume a prior for p that is equal to zero when p \< 0.5 and is a positive constant when p ≥0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.

**W, W, W**

```{r}
# define the grid, uniform prior for p

p_grid <- seq(from = 0, to = 1, length.out = 20) 

# define new prior 
prior <- ifelse(p_grid < 0.5, 0, 1)

#compute likelihood
likelihood <- dbinom(3, size = 3, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardixe Posterior 

posterior <- unstd.posterior / sum(unstd.posterior)


plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext( "3 points - WWW - new Prior " )
```

**W, W, W, L**

```{r}

# define the grid, uniform prior for p

p_grid <- seq(from = 0, to = 1, length.out = 20) # 4 observations

# define prior 
prior <- ifelse(p_grid < 0.5, 0, 1)

#compute likelihood
likelihood <- dbinom(3, size = 4, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardixe Posterior 

posterior <- unstd.posterior / sum(unstd.posterior)


plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext( "4 points - W W W L - New prior" )




```

**L, W, W, L, W, W, W**

```{r}


# define the grid, uniform prior for p

p_grid <- seq(from = 0, to = 1, length.out = 20) # 7 observations

# define prior - not sure I did this right
prior <- ifelse(p_grid < 0.5, 0, 1)

#compute likelihood
likelihood <- dbinom(4, size = 7, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardixe Posterior 

posterior <- unstd.posterior / sum(unstd.posterior)


plot( p_grid , posterior , type="b" ,
xlab="probability of water" , ylab="posterior probability" )
mtext( "7 points - L, W, W, L, W, W, W - New prior" )



```
