---
title: "Assignment 7"
author: "Madeline Berger"
date: "2026-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rstan)
options(mc.cores = parallel::detectCores()) # set up to estimate your model in parallel
rstan_options(auto_write = TRUE) # automatically save a bare version of a compiled Stan program to the hard disk so that it does not need to be recompiled

library(bayesplot)
library(dplyr)
library(ggplot2)
library(tidybayes)


```
 

50 fish were sampled and lengths were measured from 100 deep-set pelagic longlines, once in 1990 and again 2020. You want to know if there is difference in the average lengths.


### 1. Simulate data for this experiment. Assume that fish length is 1.5 times
smaller in 2020 than 1990.

```{r sim}

n <- 50
mean_1990 <- 150 # cm
mean_2020 <- 110 # cm
sd_length <- 1

# create data

data <- data.frame(
  old_new = c(rep("1_1990",n),rep("2_2020",n)),
  length = c(
    rnorm(n,mean_1990,sd_length),
    rnorm(n,mean_2020,sd_length)
  )
)


```



### 2. Fit a Stan model for your question. Discuss how you chose your priors.
Calculate the difference in means as a derived quantity.


```{r stan}

write("// Stan model for regression with categories

data {
 int < lower = 1 > N; // sample size
 vector[N] y; // response
 int x[N]; // predictor, year
 int < lower = 1 > J; // number of categories
}

parameters {
  vector[J] a; // slope(s)
  real < lower = 0 > sigma; // error
}

model {
  vector[N] mu;
  mu = a[x];
  y ~ normal(mu, sigma);
  for(j in 1:J){
    a[j] ~ normal(2,1);
  }
  sigma ~ exponential(1);
}

generated quantities {
  // difference in means
  real diff;
  diff = a[1] - a[2]; // this is subtracted posteriors. this might be wrong

  // prior predictive check;
  vector[J] a_sim;
  for(j in 1:J){
    a_sim[j] = normal_rng(2,1);
  }
  real sigma_sim = exponential_rng(1);
  real y_sim[N];
  for(i in 1:N){
    y_sim[i] = normal_rng(a_sim[x[i]], sigma_sim);  // Simulate from prior
  }
  
  // posterior predictive check;
  vector[N] y_hat;
  for (i in 1:N) {
    y_hat[i] = normal_rng(a[x[i]], sigma); // Simulated data
  }

} 
",

"stan_lm_cat.stan")

stanc("stan_lm_cat.stan")

data$X <- as.numeric(as.factor(data$old_new))
stan_data <- list(N = nrow(data), J = length(unique(data$X)), x = data$X, y = data$length)

fit_oldnew <- stan(file="stan_lm_cat.stan", data=stan_data)
fit_oldnew

```

Model checks and derived quantities: 

```{r results }

# diagnostics 
names(fit_oldnew)
mcmc_trace(fit_oldnew, pars=c("a[1]","a[2]","sigma"))


# summarize the derived difference in means

diff_in_means <- fit_oldnew %>% 
  spread_draws(diff) %>% 
  median_qi(.width = 0.5) # so this is the 50% interval that contains the difference in slopes. it croses zero?



```

The mean of the distribution of the calculated difference in means is 0.2, but its 50% interval crosses zero, indicating that there may not be a discernable difference in means of fish length between 1990 and 2020 (according to this model and my simulated data).

### 3. Plot the posterior density of the difference in the mean. How does your
simulated data compare?

```{r}

# all means calculated from data

diff_in_means_plot <- fit_oldnew %>% spread_draws(diff)


# calculated from data

data_diff <- mean(data$length[data$old_new=="1_1990" ]) - mean(data$length[data$old_new=="2_2020"])


# plot
ggplot(diff_in_means_plot, aes(x = diff)) +
         geom_density() + 
  theme_minimal()+
  geom_vline(xintercept = data_diff)


```

