---
title: "Assignment 6"
author: "Madeline Berger"
date: "2026-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE
                      )

library(rstan)
options(mc.cores = parallel::detectCores()) # set up to estimate your model in parallel
rstan_options(auto_write = TRUE) # automatically save a bare version of a compiled Stan program to the hard disk so that it does not need to be recompiled

library(bayesplot)
library(dplyr)
library(ggplot2)
library(tidybayes)
library(posterior)
library(tidyverse)
library(here)


# load the data
#library(rethinking)
data("Howell1")

d <- read_csv(here("problem_sets/Howell1.csv"))

#if ("package:rethinking" %in% search()) detach("package:rethinking", unload = TRUE)

# ------ create variables ---------- #
d$H <- d$height
d$dum <- d$male
d$index <- ifelse(d$male == 1,1,2)

```

### Note to Mary: Everything runs fine until I try to knit, after which I was getting a weird error around the "spread_draws" function. I tried to get around it multiple ways but never found a solution. So apologies that this html does not have the graphs!  


## 1. In your own words, explain the difference between the two intervals plotted in 4.10 and how you would go about creating each.

The smaller interval is the variation of mu, the mean height at each individual weight, or the posterior distribution of u. The larger interval in light grey is the interval in which you would expect 89% of the population's height would fall for each weight.

## 2. Using the Howell1 data build a model for the difference in mean height by gender using both the dummy variable and index method. Compare your results.

### Dummy Model

```{r q2, message=FALSE, warning=FALSE}




# mq.2 dummy  ------------------------------------------------------------

write("// 

data {
 int < lower = 1 > N; // sample size
 vector[N] H; // response
 vector[N] dum; // predictor
}

parameters {
  real a; // intercept
  real bM; // slope
  real sigma; // error
}

model {
  vector[N] mu;
  mu = a + bM*dum;
  H ~ normal(mu, sigma);
  a ~ normal(179,20);
  bM ~ normal(0,10);
  sigma ~ uniform(0,50);
}

generated quantities {

// prior predictive check;
  real a_sim = normal_rng(178, 20);
  real b_sim = normal_rng(0, 10);
  real sigma_sim = uniform_rng(0,20);
  real y_sim[N] = normal_rng(a_sim + b_sim * dum, sigma_sim);  // Simulate from prior

// posterior predictive check;
  vector[N] y_hat;
  for (i in 1:N) {
    y_hat[i] = normal_rng(a + bM * dum[i], sigma); // Simulated data
  }
}


",

"stan_q.1_dum.stan")

stanc("stan_q.1_dum.stan")

# fit the model 

stan_data <- list(N = nrow(d), dum = d$dum, H = d$H)

fit.dum <- rstan::stan(file="stan_q.1_dum.stan", data=stan_data, chains=3, iter=1000)
fit.dum


```

### Dummy Model Prior Checks, Diagnostics and Plots

```{r}

# Prior check - join estimation

dum_draws_array <- as.array(fit.dum)

prior_sim <- dum_draws_array %>% 
  as_draws_df() %>% 
  spread_draws(y_sim[i])

nrow(prior_sim) # 1500 iterations by 544 data points

prior_sim %>% 
  left_join(d %>% select(dum) %>% mutate(i = seq(1:544)),"i") %>% 
  mutate(dum = as.integer(dum)) %>% 
  ggplot() +
  geom_jitter(aes(x=dum,y=y_sim), alpha = 0.5, color = "skyblue", width = 0.05) +
  xlab("Gender") +
  ylab("Height")+
  theme_bw()


# diagnostics
mcmc_trace(fit.dum, pars=c("a","bM","sigma"))

# plot results
fit.dum.sum <- dum_draws_array %>% 
  posterior::as_draws_df() %>% 
  spread_draws(a, bM) %>% 
  median_qi()
d$dum_num <- as.numeric(d$dum)
y_pred <- fit.dum.sum$a+fit.dum.sum$bM*d$dum_num

plot(d$dum,d$H, col="skyblue", xlab="Gender", ylab="Height")
points(d$dum,y_pred, type = "l")

# Create prediction grid
x_grid <- data.frame(
  dum = rep(c(0, 1), each = 5)
)

# predict mu hat with data, posterior prediction
draws_with_preds <- dum_draws_array %>%
  posterior::as_draws_df() %>% 
  spread_draws(a, bM) %>% # sample from posterior, 1500
  cross_join(x_grid) %>%  # 10 samples from actual data
  mutate(mu_hat = a + bM * dum)

# get intervals from mu hat distributions
pred_intervals <- draws_with_preds %>%
  group_by(dum) %>%
  median_qi(mu_hat, .width = c(.89, .95)) 
  
# plot data, mu_hat, and interval around mu_hat  
ggplot(data = pred_intervals, aes(x = dum, y = mu_hat)) +
  geom_jitter(data=d,
              aes(x=dum,y=H),
              color="skyblue",alpha=0.4, width = 0.05)+
  geom_pointrange(aes(ymin = .lower, ymax = .upper),
                  color = "coral", size = 0.3)+
  theme_minimal()+
  labs(x = "0 = Female 1 = Male",
       y = "Height (cm)")


# if we wanted to add prediction interval for y-hat, could probably add another geom_point range?

```

### Index Model

```{r}


# mq.2 index  ------------------------------------------------------------

write("// 

data {
 int < lower = 1 > N; // sample size
 vector[N] H; // response
 int index[N]; // predictor. you dont have to add the upper and lower
}

parameters {
  vector[2] a; // two intercepts
  real < lower = 0, upper = 50 > sigma; // error
}

model {
  vector[N] mu;
  mu = a[index];
  H ~ normal(mu, sigma);
  // priors
  a[1] ~ normal(178,20); // note this is could be done with a for loop if many quantities
  a[2] ~ normal(178,20);
  sigma ~ uniform(0,50);
}

generated quantities {

// mary made a new variable for difference
  real diff;
  diff = a[1] - a[2]; // this is subtracted posteriors 
  
  
// prior predictive check
  vector[2] a_sim;
  for (i in 1:2){
   a_sim[i] = normal_rng(178,20);
  }
  real sigma_sim = uniform_rng(0,50);
  
  real y_sim[N];
  for(n in 1:N){
   y_sim[n] = normal_rng(a_sim[index[n]], sigma_sim);
  }

// posterior predictive check;
  vector[N] y_hat;
  for (n in 1:N) {
    y_hat[n] = normal_rng(a[index[n]], sigma); // Simulated data
  }
}


",

"stan_q.1_index.stan")

stanc("stan_q.1_index.stan")

# fit the model 

stan_data <- list(N = nrow(d), index = d$index, H = d$H)

fit.index <- rstan::stan(file="stan_q.1_index.stan", data=stan_data, chains=3, iter=5000)
fit.index

```


### Index Model Prior Checks, Diagnostics and Plots

```{r}

# Prior check - join estimation style

index_draws_array <- as.array(fit.index)

prior_sim <- index_draws_array %>% 
  posterior::as_draws_df() %>% 
  spread_draws(y_sim[n])

nrow(prior_sim) # 1500 iterations by 544 data points

prior_sim %>% 
  left_join(d %>% select(index) %>% mutate(n = seq(1:544)),by ="n") %>% 
  mutate(dum = as.integer(index)) %>% 
  ggplot() +
  geom_jitter(aes(x=index,y=y_sim),
             alpha = 0.5, width = 0.05, color = "aquamarine3") +
  xlab("Gender") +
  ylab("Height")+
  theme_bw() # prior y_sim range seems a bit wide but thats ok for now? 

# what params did it sample?
list_of_draws <- extract(fit.index)
print(names(list_of_draws))

# diagnostics
mcmc_trace(fit.index, pars = c("a[1]","a[2]", "sigma")) # not looking too good?

# plot results
fit.index.sum <- index_draws_array %>%
  posterior::as_draws_df() %>% 
  spread_draws(a[i],sigma) %>% 
  median_qi()
d$index_num <- as.numeric(d$index)
y_pred1 <- fit.index.sum$a[1] * d$index_num
y_pred2 <- fit.index.sum$a[2] * d$index_num # not sure this is right

plot(d$index,d$H, col="aquamarine3", xlab="Gender", ylab="Height")
points(d$index,y_pred)
points(d$index,y_pred)


# draw parameters and create mu hat predictions. adding x values based on index here instead of using cross join because you would never multiple a 2 by alpha[1], I think
draws_with_preds <- fit.index %>%
  #posterior::as_draws_df() %>% 
  spread_draws(a[i], sigma) %>% # sample from posterior, 1500 x 2 different versions of a
  mutate(index = as.numeric(i)) %>% 
  mutate(mu_hat = a*index)


# get intervals from mu hat distributions
pred_intervals <- draws_with_preds %>%
  group_by(index) %>%
  median_qi(mu_hat, .width = c(.80)) 
  
# plot data, mu_hat, and interval around mu_hat  
ggplot(data = pred_intervals, aes(x = index, y = mu_hat)) +
  geom_jitter(data=d,
              aes(x=index,y=H),
              color="aquamarine3",alpha=0.4, width = 0.05)+
  geom_pointrange(aes(ymin = .lower, ymax = .upper),
                  color = "dodgerblue", size = 0.3)+
  theme_minimal()+
  labs(x = "1 = Male, 2 = Female",
       y = "Height (cm)")


# plot density with line for 80 interval

ggplot(data = draws_with_preds, aes(x = mu_hat, group = i))+
  geom_density(stat = "density")+
  theme_minimal()
  

```

The final plot looks quite wrong but I'm not sure where I went wrong in the code. Maybe the priors? because the prior predictive plots look pretty wrong, like way too large a range for height. But I'm a bit unsure as to how the priors I specified produced such gigantic ranges. 
